---
title: "Characteristics of Adelie, Chinstrap, and Gentoo Penguins"
subtitle: "INFO 526 - Summer 2024 - Final Project"
author: 
  - name: "Michelle Chow"
    affiliations:
      - name: "School of Information, University of Arizona"
description: "Analysis of the correlation between penguin islands and bill length of Adelie, Chinstrap, and Gentoo penguins and how the body mass of penguins changed between 2007 to 2009."
format:
   html:
    code-tools: true
    code-overflow: wrap
    embed-resources: true
editor: visual
execute:
  warning: false
  echo: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)

# Load the penguin dataset
penguin <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-04-15/penguins.csv')

# Set theme for ggplot2
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 14))

# Set width of code output
options(width = 65)

# Set figure parameters for knitr
knitr::opts_chunk$set(
fig.width = 7, # 7" width
fig.asp = 0.618, # golden ratio
fig.retina = 3,
fig.align = "center", # center align figures
dpi = 300 
)
```

## Abstract

The project will explore the how different characteristics of penguins differ between different islands and over time. The Palmer penguin dataset, which was first created in TidyTuesday in July 2020 and most recently updated in 2025, will be used to carry out the analysis. The dataset covers 3 species of penguins - Adelie, Chinstrap, and Gentoo, with 344 observations that cover 5 different penguin characteristics - Bill length, Bill depth, Flipper length, Body Mass, and Sex between 2007 and 2009. The analysis will specifically focus on using the variables, Bill length, Body mass, Species, Island, and Year to answer two questions. The first question will explore the correlation between the bill length of Adelie, Gentoo, and Chinstrap penguins and the island in which they reside upon. The second question will then explore how the body mass of penguins have changed between 2007 and 2009.

## Introduction

Penguins are flightless birds that are highly specialized in swimming and diving, replacing wings with flippers. They excel in water and can survive freezing temperatures. Although many species share common features, specific characteristics, such as size, bill length, body mass etc., can still differ. These characteristics have changed over time to adapt to their habitat and can play a vital role in determining a penguins feeding strategy, diet, and thermoregulation. Through analysis, we can gain insight into how the specific penguin species, Adelie, Gentoo and Chinstrap, have adapted to their habitat over time.

## Question 1: How does the penguin species and the island where Adelie, Gentoo, and Chinstrap penguins reside impact their bill length?

### Introduction

The bill of penguins is critical to their survival. Penguins utilize their beaks to hunt and catch their prey and regulate their body temperature. Therefore, it is important to understand how the length of their bill has adapted to the islands in which they reside. This question aims to answer whether bill lengths vary based on the species of the penguin, and if so, if the island of habitation can additionally have an impact on the bill length.

### Approach

To answer this question, a bar chart will be created to first understand which islands each penguin resides. This plot will show the number of Adelie, Chinstrap, and Gentoo penguins on each island. We can then use this data to determine which species can be used to compare the impact of islands on bill length.

Then, a box plot will be created using the `bill_len`, `species`, and `island` variable to compare the median and spread of bill lengths of Adelie, Chinstrap, and Gentoo penguins. The box plot will also provide insight into how the median and spread of bill lengths differ between islands for each species. From this, we can deduce which species of penguins usually have longer or shorter bills and whether their habitat plays a role on the length of the bill.

Next, a violin plot will be created to better look at the distribution of bill lengths for each species and how they differ across islands. The plot will also be able to show if bill lengths are concentrated at a certain length to show whether the data is skewed.

### Analysis

```{r}
# Create a bar plot showing which islands the Gentoo, Adelie, and Chinstrap penguins inhabit
ggplot(penguin, aes(x = island, fill = species)) +
  geom_bar(position = "dodge") +
  labs(
    x = "Island",
    y = "No. of Penguins",
    fill = "Penguin Species"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

The bar chart showed that, although the dataset contained 3 different islands, only Adelie penguins inhabited all 3 islands. Gentoo penguins were concentrated on Biscoe, while Chinstrap penguins resided on Dream. Therefore, based on this graph, we can only compare the impact of the island of habitation on bill length for Adelie penguins. This chart also shows that Adelie and Gentoo penguins both had similar number of observations (\~120 observations) while there were only approximately 60 observations for Chinstrap penguins. Based on this graph we can then use a box plot to analyze the difference in bill length between Adelie, Chinstrap, and Gentoo penguins and how living on a different island impacts an Adelie's bill length.

```{r}
# Create a box plot of Gentoo, Adelie, and Chinstrap penguin bill lengths
ggplot(penguin, aes(x = species, y = bill_len, fill = island)) +
  geom_boxplot() +
  labs(
    x = "Penguin Species",
    y = "Bill length (mm)",
    title = "Bill length of Adelie, Chinstrap, and Gentoo Penguins",
    fill = "Island"
  ) +
  guides(fill = guide_legend(override.aes = list(shape = 22, size = 5, color = NA))) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

A box plot was then created to show how the bill length differed between penguin species and islands. Chinstrap penguins had the highest median bill length (\~49mm) relative to Gentoo and Adelie penguins. Gentoo penguins had a bill length of approximately 47mm while Adelie penguins had the shortest bills, with a median bill length of approximately 38mm. Therefore, it can be deduced that chinstrap penguins usually have longer bills compared to Gentoo and Adelie penguins. However, there was one outlier for Gentoo penguins, with a bill length of 60mm. This Gentoo penguin had the longest bill length of all the penguins in the dataset. Adelie penguins had similar median bill lengths for each of the 3 islands. However, Adelie penguins residing in Biscoe, had the least spread/difference in bill lengths whilst Torgersen and Dream Adelie penguins had a larger difference in bill length between each penguin. The box plot also shows that Chinstrap penguins had the largest spread in bill lengths amongst all the penguin species, suggesting a large variation in bill length between different Chinstrap penguins.

```{r}
# Create a violin plot of the distribution of bill lengths for each of the penguin species
ggplot(penguin, aes(x = species, y = bill_len, fill = island)) +
  geom_violin() +
  labs(
    x = "Penguin Species",
    y = "Bill length (mm)",
    fill = "Island",
    title = "Bill length of Adelie, Chinstrap, and Gentoo Penguins"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

Lastly, the violin plot was created to provide insight into the distribution of the bill lengths for Adelie, Chinstrap, and Gentoo penguins. From the plot, it is evident that most Chinstrap penguins have a bill length of approximately 51mm, with a moderate amount of Chinstrap penguins with a bill length of \~46mm. Only a small amount of Chinstrap penguins had a bill length greater than 52mm and shorter than 45mm due to the narrow width of the distribution. Gentoo penguins, on the other hand, usually have a bill length of 45mm - 50mm as seen by the wide section. The data is right-skewed as there are only a few observations that have a bill length larger than \~51mm and the majority of the observations were concentrated near the bottom. As for Adelie penguins, the have bill lengths between \~37mm and \~41mm. The bill length for Adelie penguins on Biscoe and Torgersen, tend to be longer than those living on Dream. The distribution of the bill lengths for the Torgersen island are relatively more symmetrical compared to Biscoe and Dream which are right-skewed and left-skewed respectively.

### Conclusion

In conclusion, based on the box and violin plot, Chinstrap penguins are often correlated with longer bill lengths than Gentoo and Adelie penguins. Adelie penguins, on the other hand, more commonly have shorter bill lengths than the other 2 species. Chinstrap penguins are also more likely to have larger differences in bill length, varying from \~41mm to \~58mm, however, the violin plot shows that only a few penguins were observed to have bill lengths larger than \~51mm with the majority having bill lengths between 45 - 50mm. Adelie penguins, especially those residing on Biscoe, have the least difference in bill lengths.

As for the impact of the island of habitation on bill length, Adelie penguins were seen to have the most evenly spread bill lengths when residing on Torgersen, with a large range in length. Adelie penguins on Dream, on the hand, tended to have shorter bills than the other 2 islands.

## Question 2: How has the body mass of the Adelie, Gentoo, and Chinstrap penguins changed between 2007 and 2009 and does the island impact the mass?

### Introduction

Body mass is also another critical factor to a penguins survival. It can impact a penguins chance of survival, reproduction, and thermal insulation. As a result, it is important to track the body mass of penguins to ensure there are no concerning changes in a penguin colony. Furthermore, by comparing data from previous years, a baseline can be drawn to show changes over the years. This question aims to answer whether there has been any changes to body mass for Adelie, Gentoo, and Chinstrap penguins, and if so, how and when.

### Approach

To answer this question, the mean body mass of Adelie, Gentoo, and Chinstrap penguins will be calculated for 2007, 2008, and 2009 and placed into the variable `penguin_mean`. Once the mean body mass is calculated for every year and species, a time series plot will then be created so that the mean body mass of the 3 different species can be compared.

Following the time series plot, another time series plot will then be made. As the Adelie penguin is the only penguin that resides in more than 1 island, it will be used to conduct the analysis on whether the habitat area plays a role on the mean body mass of Adelie penguins. The plot will compare Adelie penguins and their mean body mass based on their habitat island between 2007 and 2009.

### Analysis

```{r}
# Calculate the mean body mass of the Adelie, Gentoo, and Chinstrap penguin for 2007, 2008, and 2009
penguin_mean <- penguin |>
  group_by(year, species) |>
  summarize(
    mean_body_mass = mean(body_mass, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r}
# Create a time series plot to compare the mean body mass of Adelie, Gentoo, and Chinstrap penguins between 2007 and 2009
ggplot(
  penguin_mean, 
  aes(x = factor(year), y = mean_body_mass, color = species, group = species)) +
    geom_line() +
    geom_point() +
    labs(
      x = "Year",
      y = "Mean Body Mass (g)",
      color = "Species",
      title = "Mean Body Mass of Adelie, Chinstrap, and Gentoo penguins between 2007 and 2009"
    ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

The time series plot shows that out of the 3 species, Gentoo penguins have the highest mean body mass (\~5100g) while Adelie and Chinstrap penguins have similar mean body mass (\~3700g). Gentoo penguins saw an approximate 100g drop in their mean body mass in 2008 which then recovered and increased to approximately 5150g in 2009, leading to a 50g net change to their mean body mass. Adelie and Chinstrap penguins, on the other hand, had the reverse trend. These species saw an uptick in their mean body mass in 2008 before dropping again in 2009. Specifically, the Adelie saw the least changes in their body mass - Increasing by approx. 50g in 2008 before decreasing to 3650g in 2009, with a net change of \~50g. Chinstrap penguins followed a similar trend but had larger fluctuations in their mean body mass. The mean body mass increased by 100g in 2008 before decreasing by 100g in 2009 back to their original weight.

```{r}
# Calculate the Adelie penguins' mean body mass between 2007 and 2009 for each of the 3 islands
adelie_mean <- penguin |>
  filter(species == "Adelie") |>
  group_by(year, island) |>
  summarize(
    mean_body_mass = mean(body_mass, na.rm = TRUE),
  )
```

```{r}
# Create a time series plot to compare the mean body mass of Adelie penguins between 2007 and 2009 for each of the 3 islands
ggplot(
  adelie_mean,
  aes(x = factor(year),
  y = mean_body_mass,
  color = island,
  group = island
  )
) +
  geom_line() +
  geom_point() +
  labs(
    title = "Mean Body Mass of Adelie penguins between 2007 and 2009",
    x = "Year",
    y = "Mean Body Mass (g)",
    color = "Island"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(size = 10, hjust = 0.5))
```

Based on the time series plot, there are big fluctuations in the Adelie penguin mean body mass between 2007 and 2009, especially those who reside on the Torgersen and Biscoe Islands. Specifically, on the Torgersen island, Adelie penguins saw an increase in mean body mass by of approximately 100g to 3850g in 2008. However, the mean body mass drops to less than 3500g in 2009. The dramatic decrease can be a cause for concern. On the Biscoe island, Adelie penguins saw a exact opposite trend. The mean body mass first increases a little (\~10g) in 2008 before a steep increase to 3850g in 2009. As for Dream, Adelie penguins had an moderate increase in mean body mass in 2008 (from 3675g to 3750g) before dropping back to similar original levels in 2009 (from 3750g to 3650g).

### Conclusion

In conclusion, Gentoo penguins have the highest mean body mass (\~5100g) of the 3 penguin species analyzed while Adelie and Chinstrap penguins had similar mean body masses (\~3700g). Gentoo penguins also had a net positive change in their mean body mass between 2007 and 2009 (+50g) while Adelie and Chinstrap penguins saw a drop to their mean body mass (-50g and +0g, respectively).

As for Adelie penguins, specifically, Adelie penguins on Biscoe had a increase in their mean body mass between 2007 to 2009 (from 3625g to 3850g). Adelie penguins on Dream and Torgersen, on the other hand, had a decrease in their mean body mass. Those on the Torgersen Island faced the steepest drop from 3755g to less than 3500g (net decrease of 255g) while Adelie penguins living on Dream had a drop from 3675g to 3650g (net decrease of 25g). The steep drop faced by Adelie penguins may be a cause for concern, however, more data needs to be gathered to form a clearer correlation between time and mean body mass.