---
title: "Palmer Penguins Analysis"
subtitle: "Proposal"
author: 
  - name: "Michelle Chow"
    affiliations:
      - name: "School of Information, University of Arizona"
description: "This project analyzes the impact of islands on the bill length of Adelie, Chinstrap, and Gentoo penguins. It also compares how the body mass of these penguins change between 2007 to 2009."
format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: true
    embed-resources: true
editor: visual
code-annotations: hover
execute:
  warning: false
---

```{r}
# Load required packages
library(tidyverse)
```

## Dataset

```{r}
# load the dataset
penguins <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-04-15/penguins.csv')
penguins_raw <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-04-15/penguins_raw.csv')
```

```{r}
# View the first 10 lines of the dataset
head(penguins)
head(penguins_raw)
# Preview the unique penguin species in the dataset
unique(penguins$species)
# Preview the unique islands in the dataset
unique(penguins$island)
```

## Brief description of the dataset

The Palmer Penguins dataset was first created in TidyTuesday in July 2020 and was last updated in 2025. The dataset has 8 columns/variables - Species, Island, Bill length, Bill depth, Flipper length, Body mass, Sex, and Year. There are 344 observations/rows of data on the various species of penguins. The species included in this dataset includes 3 types of penguins - Adelie, Gentoo, and Chinstrap. The dataset also focuses on 3 islands - Torgersen, Biscoe, and Dream. I chose this dataset as I love penguins and I would love to see how different factors can affect the different characteristics in penguins. It would also be interesting to see how the body mass of penguins change over the years through evolution.

## Questions

1.  How does the island where Adelie, Gentoo, and Chinstrap penguins reside impact their bill length?
2.  How has the body mass of the Adelie, Gentoo, and Chinstrap penguins changed between 2007 and 2009?

## Analysis plan

- A plan for answering each of the questions including the variables involved, variables to be created (if any), external data to be merged in (if any).

**Question 1 plan**

Variables involved:

\- Island the penguin resides (Torgersen, Biscoe, Dream)

\- Species of penguins (Adelie, Gentoo, Chinstrap)

\- Bill length (mm)

**Plan:**

1.  Create side-by-side box plots to compare bill lengths of penguin species and their bill length

Box plot variables:

x-axis = Species of penguins (Adelie, Gentoo, Chinstrap)

y-axis = Bill length (mm)

fill = Island (Torgersen, Biscoe, Dream)

Create a violin plot to compare the distribution of bill lengths for each penguin species

x-axis = Species of penguins (Adelie, Gentoo, Chinstrap)

y-axis = Bill length (mm)

fill = Island (Torgersen, Biscoe, Dream)

**Question 2 plan**

Variables involved:

\- Body mass of penguins (g)

\- Penguin species (Adelie, Gentoo, Chinstrap)

\- Year (2007, 2008, 2009)

**Plan:**

Create a time series plot to compare the body mass of penguins for each species between 2007 and 2009

Time series plot variables:

x-axis = Year (2007, 2008, 2009)

y-axis = Body mass of penguins (g)

fill = Penguin species (Adelie, Gentoo, Chinstrap)

## Weekly Project Plan Table

| Task Name | Status | Due | Priority | Summary |
|---------------|---------------|---------------|---------------|---------------|
| Create presentation | Not completed | 28-06-2026 | Low | Create 7 presentation slides and create recording of the presentation |
| Add interpretation of the plot | Not completed | 21-06-2026 | Moderate | Discuss the results of the time series, violin, and box plots |
| Write introduction and approach | Not completed | 21-06-2026 | Moderate | Create introduction of the dataset and approach |
| Create time series, violin, and box plots | Not completed | 17-06-2026 | High | Create the plots for analysis using the dataset |