---
title: "Proposal title"
subtitle: "Proposal"
author: 
  - name: "Michelle Chow"
    affiliations:
      - name: "School of Information, University of Arizona"
description: "Project description"
format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: true
    embed-resources: true
editor: visual
code-annotations: hover
execute:
  warning: false
---

```{r}
#| label: load-pkgs
#| message: false
library(tidyverse)
```

## Dataset

```{r}
# load the dataset
penguins <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-04-15/penguins.csv')
penguins_raw <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-04-15/penguins_raw.csv')
```

```{r}
# View the first 10 lines of the dataset
head(penguins)
head(penguins_raw)
unique(penguins$species)
```

## A brief description of your dataset including its provenance, dimensions, etc. as well as the reason why you chose this dataset.

## Make sure to load the data and use inline code for some of this information.

The Palmer Penguins dataset was created in TidyTuesday in July 2020. The dataset has 8 columns/variables - Species, Island, Bill length, Bill depth, Flipper length, Body mass, Sex, and Year. There are 344 observations/rows of data on the various species of penguins. The species included in this dataset includes 3 types of penguins - Adelie, Gentoo, and Chinstrap. I chose this dataset as I love penguins and I would love to see how different factors can affect the different characteristics in penguins. It would also be interesting to see how the body mass of penguins change over the years through evolution.

## Questions

1.  How does the island where the penguin resides affect the bill length?
2.  How has the body mass of the different species of penguins changed over the years?

## Analysis plan

- A plan for answering each of the questions including the variables involved, variables to be created (if any), external data to be merged in (if any).

**Question 1 plan**

Variables involved:

\- Island the penguin resides

\- Species of penguins

\- Bill length

**Plan:**

1.  Create side-by-side box plots to compare bill lengths of penguin species and their bill length

Box plot variables:

x-axis = Species

y-axis = Bill length

fill = Island

Create a violin plot to compare the distribution of bill lengths for each penguin species

x-axis = Species

y-axis = Bill length(mm)

fill = Island

**Question 2 plan**

Variables involved:

\- Body mass of penguins

\- Penguin species

\- Year

**Plan:**

Create a time series plot to compare the body mass of penguins for each species over the years

Time series plot variables:

x-axis = Year

y-axis = Body mass of penguins

fill = Penguin species

## Weekly Project Plan Table Task Name:

| Task Name | Status | Due | Priority | Summary |
|---------------|---------------|---------------|---------------|---------------|
| Create presentation | Not completed | 28-06-2026 | Low | Create 7 presentation slides and create recording of the presentation |
| Add interpretation of the plot | Not completed | 21-06-2026 | Moderate | Discuss the results of the time series, violin, and box plots |
| Write introduction and approach | Not completed | 21-06-2026 | Moderate | Create introduction of the dataset and approach |
| Create time series, violin, and box plots | Not completed | 17-06-2026 | High | Create the plots for analysis using the dataset |